Manuscript draft

### 2019-06-05 manuscript/manuscript draft.Rmd
updated manuscript draft

add hypotheses, can number/bullet point them
redo first couple paragraphs
look again at breakdown of primary productivity components --> add discussion on understanding belowground productivity

@@

There are two major questions to understand: firstly, how forest productivity varies globally, and specifically which areas show the greatest peaks in productivity; and secondly, which climate variables drive this variation. On a global scale, the productivity of forests varies with latitude, with a general trend of increasing productivity towards the tropics [@Beer2010; @Jung2011]; however the exact nature of this pattern, and how it varies by component of productivity, is poorly understood. This latitudinal gradient is most likely to be explained by climatic gradients in temperature, precipitation, length of growing season, and combinations of the above. Productivity is influenced by a range of climatic drivers, including mean annual temperature (MAT) and mean annual precipitation (MAP) [@DelGrosso2008], but doesn't necessarily respond linearly to these drivers. Disentangling the shape of productivity responses to climate drivers will enable better predictions of future responses under climate change.

*Patterns of primary productivity.* Understanding variation in primary productivity is complex. Firstly, there are many factors which can influence productivity on a range of scales, many of which show interactive effects with each other. In addition, the primary productivity of a forest is the sum of the primary productivity of a number of individual components, which may all show individual responses to drivers of productivity. Understanding these relationships and interdependencies is complex.

Primary productivity of forests is influenced by a range of factors. On a local scale, stand age [@Litton2007; @Gillman2015], management [@Simova2017]; nutrient availability [@Aragao2009]; and altitude [@Girardin2010; @Malhi2017] all impact stand productivity. On a global scale, changes in primary productivity are influenced by climatic variables and abiotic gradients, such as the length of growing season [@Michaletz2014]. The majority of studies to date have focused on the influence of MAP and MAT on productivity. While MAP and MAT have been argued to be significant predictors of productivity [@Chu2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@Kerkhoff2005; @Malhi2012; @Michaletz2014; @Michaletz2018]. The influence of other climate variables on global patterns of productivity remains under-explored.

*Understanding components of primary productivity.* There are several different measures of primary productivity. Gross primary productivity (GPP) represents the gross uptake of carbon dioxide by plants in an ecosystem, to be used in photosynthesis and energy production. Only a fraction of this carbon is assimilated into plant tissues; the rest is used in autotrophic respiration. The component of GPP that is stored as plant material is the net primary productivity (NPP) of an ecosystem; thus GPP can be described as the sum of NPP and ecosystem respiration.   
Currently, GPP cannot be measured directly by observing total ecosystem photosynthesis. Instead field estimates of GPP have to be derived based on modelling and extrapolation of eddy-covariance studies and measurements of net ecosystem exchange (NEE) [@Clark2017].

NPP is typically measured through direct measurements of the productivity of its components, which are then summed to generate a whole-ecosystem estimate. These components can be broadly divided into aboveground NPP (ANPP) and belowground NPP (BNPP). Measurement of aboveground NPP - primarily consisting of woody stem, woody branch and canopy/foliage primary productivity, but also sometimes including estimates of productivity of reproductive and understory components, and losses due to herbivory - is most commonly estimated through forest plot censuses. Belowground NPP is significantly more challenging to measure, meaning that estimates of belowground productivity are less reliable than those of aboveground productivity. Coarse root productivity is typically determined by using allometric equations relating root mass to stem diameter. This relies on extrapolation of ANPP measures, rather than being based on direct measures of root productivity. Fine root productivity is estimated using a variety of methods, such as taking soil cores and minirhizotrons, making it typically more reliable than measures of coarse root productivity.

[[Belowground productivity is a potentially often underestimated component of productivity, but has been shown to be significant, particularly in higher latitude forests [@Pan2011].]]

Current knowledge indicates that these different components of productivity show different latitudinal patterns. Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@Beer2010; @Jung2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and mean annual temperature (MAT) and mean annual precipitation (MAP) [@Luyssaert2007], with the highest GPP values reported in tropical forests. The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@Luyssaert2007; @Huston2009]. Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. [-@Gillman2015] found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@Huston2009].

These variations in the responses of different components of productivity to changes in latitude suggest that components may vary in the strength of their response to climatic drivers. There is some empirical evidence for this: increases in GPP are reported to saturate above 25$^\circ$C MAT [@Larjavaara2012], while increases in NPP are recorded to saturate above 10$^\circ$C MAT [@Luyssaert2007]. Similarly, allocation to different components of ANPP varies with climate. Within the tropics, allocation to canopy NPP appears fairly consistent, with significantly greater variation in allocation to woody and belowground NPP [@Litton2007; @Malhi2011]. Allocation to these structural biomass components has been shown to increase with water availability [@Litton2007; @Bloom2016], and is highest in the wet tropics, indicating that control of woody productivity by MAP may be more significant than control of foliar productivity. However, these studies are regional, meaning that our understanding of variation in allocation and its relationship to climate on a global scale remains limited.

Allocation is also influenced by stand age [@DeLucia2007], nutrient availability [@Litton2007] and forest structure [@Taylor2019], which can make it challenging to disentangle the effects of climate.

Data that control for stand age and standardize methodologies are required to resolve this question. Here, we use a comprehensive global database to explore global patterns in productivity. We explore three questions: 
1. Which climatic variables are the most important drivers behind the latitudinal pattern in primary productivity?
2. Do the components of primary productivity show variation in their responses to these climatic drivers? Is carbon allocation climate invariant?
3. Does climate explain the same proportion of variation in different components of primary productivity? Is the relative importance of climatic drivers the same across components of primary productivity?

@@

---
output:
  pdf_document:
    fig_caption: yes
---

![Figure 1: Map showing all data used in the anlysis, coded by variable \label{figurelabel}](distribution_all_variables.png)

@@

### References

>>>>

, fig.cap = "Map showing all data used in the analysis, coded by variable"

@@

There are two major questions to understand: firstly, how forest productivity varies globally - and specifically which areas show the greatest peaks in productivity -; and secondly, which climate variables drive this variation. On a global scale, the productivity of forests varies with latitude, with a general trend of increasing productivity towards the tropics [@Beer2010; @Jung2011]; however the exact nature of this pattern, and how it varies by component of productivity, is poorly understood. This latitudinal gradient is most likely to be explained by climatic gradients in temperature, precipitation, length of growing season, and combinations of the above. Productivity is influenced by a range of climatic drivers, including mean annual temperature (MAT) and mean annual precipitation (MAP) [@DelGrosso2008], but doesn't necessarily respond linearly to these drivers. Disentangling the shape of productivity responses to climate drivers will enable better predictions of future responses under climate change.

*What is primary productivity?* During photosynthesis, plants capture carbon dioxide from the atmosphere. The gross primary productivity (GPP) of an ecosystem is the gross uptake, via photosynthesis, of carbon dioxide by plants in that ecosystem. Only a fraction of the carbon captured is assimilated into plant tissue; the rest is used in autotrophic respiration (R~a~). The component of GPP that is stored as plant material is the net primary productivity (NPP) of an ecosystem. Net primary productivity can therefore be expressed as:

$$ NPP = GPP + R~a~ $$

Currently, GPP cannot be measured directly by observing total ecosystem photosynthesis. Instead field estimates of GPP have to be derived based on modelling and extrapolation of eddy-covariance studies and measurements of net ecosystem exchange (NEE) [@Clark2017].

In contrast, NPP can be calculated through direct field observations. In order to achieve greater accuracy in estimating NPP, NPP is often broken down into its component parts, with aboveground NPP (ANPP) and belowground NPP (BNPP) considered separately to each other. The components included when estimating ANPP and BNPP often vary between studies, depending on the intensity of fieldwork effort. At its most basic level, ANPP can be expressed as:

$$ ANPP = NPP~stem~ + NPP~branch~ + NPP~canopy~ $$
where NPP~stem~ is the annual woody increment of all stems above a specified diameter at breast height (DBH), NPP~branch~ is annual woody branch turnover, and NPP~canopy~ is annual foliage production, including leaves, twigs, and reproductive structures. ANPP may also include NPP~VOC~, the annual emission of volatile organic compounds, and NPP~herbivory~, the annual consumption of plant matter by herbivores, but these components are often excluded from field observations as they are much harder to quantify. Other components of aboveground productivity that remain largely unquantified include epiphytes, hemiepiphytes, and understory plants [@Clark2017]. All current ANPP estimates are based on the assumption that the contribution of these components to overall NPP is insignificant.

There are two major subcomponents of BNPP, which can be expressed as:

$$ BNPP = NPP~coarse root~ + NPP~fine root~ $$
where NPP~coarse root~ is the annual production of coarse roots (typically roots >2mm diameter), and NPP~fine root~ is the annual production of fine roots (typically roots <2mm diameter) [@Aragao2009]. Calculations of BNPP may also include NPP~exudation~, a measure of annual carbon losses through root exudation, and NPP~symbionts~, the annual carbon allocation to mycorrhizae and legumes, but, as before, this is challenging to quantify and is often excluded from field observations.

BNPP is a poorly understood component of total ecosystem productivity, primarily because of the challenges in gaining accurate field measurements. Coarse root productivity is often estimated via extrapolation of NPP~stem~ estimates using allometries that may not have been empirically verified [@Clark2001]. NPP~fine root~ is easier to quantify through soil cores and minirhizotrons, however, sampling tends to be limited to the surface soils, with very few studies sampling to depths below 3 metres [cite]. As a result, it is possible that BNPP is currently significantly underestimated, despite being a hugely significant component of total ecosystem productivity [@Pan2011].

*Which factors influence primary productivity?* Primary productivity is influenced by many factors, which often act across a range of scales, and may show interactive effects with each other. On a local scale, stand age [@Litton2007; @Gillman2015], management [@Simova2017]; nutrient availability [@Aragao2009]; and altitude [@Girardin2010; @Malhi2017] all impact forest productivity. On a global scale, changes in primary productivity are influenced by climatic variables and abiotic gradients, such as the length of growing season [@Michaletz2014]. There is some debate over the precise relationship between these drivers and productivity; While mean annual precipitation (MAP) and mean annual temperature (MAT) have been argued to be significant predictors of productivity [@Chu2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@Kerkhoff2005; @Malhi2012; @Michaletz2014; @Michaletz2018]. Improving understanding of how these factors interact to control global patterns in primary productivity is essential to understanding the global carbon cycle.

Current research into how primary productivity varies with latitude is inconclusive, and - though it has primarily focussed on patterns of GPP, NPP, and ANPP - indicates that different components of productivity may show different relationships to latitude. Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@Beer2010; @Jung2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and MAT and MAP [@Luyssaert2007], with the highest GPP values reported in tropical forests. The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@Luyssaert2007; @Huston2009]. Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. [-@Gillman2015] found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@Huston2009].

Furthermore, there is evidence that different components of productivity show individual responses to drivers of productivity. For example, increases in GPP have been reported to saturate above 25$^\circ$C MAT [@Larjavaara2012], while increases in NPP are recorded to saturate above 10$^\circ$C MAT [@Luyssaert2007]. Similarly, allocation to different components of ANPP varies with climate. Within the tropics, allocation to canopy NPP appears fairly consistent, with significantly greater variation in allocation to woody and belowground NPP [@Litton2007; @Malhi2011]. Allocation to these structural biomass components has been shown to increase with water availability [@Litton2007; @Bloom2016], and is highest in the wet tropics, indicating that control of woody productivity by MAP may be more significant than control of foliar productivity. However, these studies are regional, meaning that our understanding of variation in allocation and its relationship to climate on a global scale remains limited. In addition, allocation is also influenced by stand age [@DeLucia2007], nutrient availability [@Litton2007] and forest structure [@Taylor2019], which can make it challenging to disentangle the effects of climate.

Data that control for stand age and standardize methodologies are required to resolve this question. Here, we use a comprehensive global database to explore global patterns in productivity. We explore three questions: 
1. **Which climatic variables are the most important drivers behind the latitudinal pattern in primary productivity?** To date, the majority of studies have focussed on productivity responses to precipitation and temperature, while the influence of other climate variables on productivity remains under-explored. We utilise global climate datasets to investigate the relationships between productivity and a range of climate variables.  
2. **Do the components of primary productivity show variation in their responses to these climatic drivers?** Allocation to components of primary productivity appears to show some variation across climate and latitudinal gradients. We use global datasets of a range of primary productivity components to explore how significant this variation in allocation is.  
3. **Does climate explain the same proportion of variation in different components of primary productivity?** Review of current research shows that, while the relationship between GPP and latitude is consistent, NPP and ANPP show weaker relationships with latitude. This may be a result of varying sensitivities to climate drivers. We use mixed models to estimate the proportion of variation explained by climate for different components of productivity, to investigate how the relative importance of climatic drivers varies across components.

@@

  ```{r echo=FALSE, out.width='100%'}
  knitr::include_graphics("distribution_all_variables.png")
  ```

@@
